Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-06 paper/ms.tex
update ms

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of FGK stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
  past 3\,Gyr, equally for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is connected to the star formation
  history of the Galaxy, and it helps clarify the relative balance of
  young and old planets that have so far been detected using transits.
  %
\end{abstract}

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
rarity of young stars: under these assumptions, $\approx$1\% of nearby
stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
Studies of currently forming protoplanets \citep{2018A&A...617A..44K},
or of exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the Galactic star
formation rate.

Despite the observational challenges, young close-in planet discovery
has done well over the past decade, primarily due to
by Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar ages are
precise at the $\approx$10\% level, and planets discovered in this way
are generally assumed to have the same age as the host cluster.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

The main problem with the ``cluster-first'' approach is that most
nearby sub-gigayear stars are in the field.  Only $\approx$1\% of
stars within the nearest $\approx$500\,pc are currently associated
with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing reasonable precision requirements\footnote{For
instance $t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
S/N$=$3.} gives $\approx$50 ``young'' transiting planets at the time
of writing.  Most of these planets are in clusters, and have been
discovered by Kepler, K2, or TESS.  However, combined, those surveys
have discovered $\approx$5{,}000 planets.  Unless there is a major
bias against the discovery of sub-gigayear old planets with precisely
constrained ages, we would expect an order of magnitude more young
transiting planets to be known.

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for stars in the galactic thin
disk, particularly young stars?  Second, where are the missing young
transiting planets?  Briefly, we will find that uniform is incorrect
by a factor of a few, and that the bias against the discovery of young
transiting plantes has, to date, primarily been in the measurement of
theirs ages, rather than in the sensitivity of the transit searches.

Given that Kepler's primary mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged both isochronal ages from
an earlier analogous study \citep{Fulton_Petigura_2018_cks_vii}, and
used a relative gyrochrone age-ranking approach that sorted stars
based on whether they rotated faster or slower than a gyrochrone age
of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
were conducted in the broader context of investigating trends in
exoplanet demographics, which also depend on stellar metallicity and
mass \citep[e.g.][]{Petigura_2018}.  These studies offered limited
sensitivity to the ages of the youngest planets, due to the nature of
isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, as well
    as our understanding of which of these objects are false
    positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    However, for late G dwarfs and K dwarfs, the age-calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 that comprised
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}; the
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  The logarithmic scale highlights the
    dynamic range of the detections and upper limits.  Some stars with
    lithium detections do not have detected rotation periods, and
    vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

@@

Our EW measurement approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
precision of our equivalent measurement technique, after applying an
initial iteration on the \nlithiumstars\ KOIs with spectra, we
compared our lithium equivalent widths with those reported by
\citet{2018ApJ...855..115B}.  For the \nbergeroverlap\ stars in both
samples, we found broad agreement at $>$30\,m\AA, and significant
differences at $<$30\,m\AA because \citet{2018ApJ...855..115B}
required positive EWs, while we allow for negative ones.  At
$>$30\,m\AA, there is however a small offset in our respective
scales, such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  This offset is
caused by our non-treatment of the iron blend.  We therefore directly
subtract the constant mean value noted above when calculating
lithium-based ages.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun}, last
accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
because 91 KOIs are omitted in the 1$''$ match, 23 of which have
either ``candidate'' or ``confirmed'' status.  Almost all of these 23
are M dwarfs with high proper motions.  Performing the same crossmatch
at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
false positives, fail to yield Gaia DR3 crossmatches.}, which is a
crossmatch of the NASA Exoplanet Archive
\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ search radius, we adopted the brightest star
as the default match.  In most such cases this was unambiguous because
there is a large brightness difference between the primary and any
apparent neighbors.  However cases with multiple stars within 4$''$
within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This is a
challenge for rotation-based age-dating because {\it i)} point sources
far from the main-sequence include unresolved photometric binaries,
and {\it ii}) effects other than magnetic braking influence the
rotation periods of bona fide single stars far from the main-sequence.
The added complications of reddening and metallicity make defining a
simple cut in any two dimensions challenging.  
We explored various options, and settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.
While the details of how exactly this locus is constructed are
arbitrary and convoluted,\footnote{In practice, for K dwarfs we fitted an
$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}. We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  For F and G dwarfs, this procedure failed to produce
a reasonable locus because of shifts in the relative precision of
the isochrone ages.  We therefore visually selected KOIs that appeared
to be near the ``main sequence'', fitted a separate polynomial to
their locations, and then stitched the resulting polynomials from the
F and G dwarfs to that of the K dwarfs.} the general 
aim is to exclude stars for which there is photometric evidence
that gyrochronology may not be reliable.  The rotation period
distribution of sources excluded in this manner includes a relatively
larger fraction of ``rapid rotators'' than the main sequence sources,
likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or Figures~12 and 13
of \citet{2023ApJS..268....4F}.

@@

\subsection{Planet Quality Flags}
\label{subsec:plflags}
Not all planets are born equal.
We use the following additional quality indicators to assess 
the reliability and utility of a planet, and assemble them into a separate bit-flag,
$Q_{\rm planet}$.

@@

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Kepler's demographic cliff,} visible in the rotation-derived
    ages of stars (left) and planet hosts (right) in the Kepler field.
    The lightly colored histograms correspond to all
    \nuniqstarsantosrotteffcut\ $T_{\rm eff}\in[3800,6200]\,$K Kepler
    stars with reported rotation periods, and all
    \nplhoststarwgyroage\ planet-hosting stars with reported rotation
    periods.
    The darker histograms impose the necessary quality cuts on factors
    such as binarity, crowding, and the star's evolutionary state
    (Section~\ref{subsec:flags}).
    In detail, these ``histograms'' are resampled from the posterior
    probability distributions, and so have less resolution at
    $\lesssim$0.5\,Gyr, when the relative precision of rotation-based
    ages decreases.
    Completeness is near unity at t$\lesssim$3\,Gyr
    \citep{2022ApJ...937...94M}.
    {\bf todo: numbers dont match.  why?}
    \label{fig:hist_tgyro}
  }
\end{figure*}

@@

Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
stars and planet-hosts in the Kepler field.  We truncated the plots
near $3$\,Gyr due to arguments that suggest the detectability of
rotation signals is near-unity up to this age for Sun-like stars, and
decreases for older ages \citep{2022ApJ...937...94M}.  Both the
stellar and planetary samples point toward a star formation rate in
the Kepler field that has slowed over the past few gigayears.

Imagine we were to group stars between 0-1\,Gyr, 1-2\,Gyr, and
2-3\,Gyr, and label them ``young'', ``intermediate-age'', and ``old''.
A simple counting exercise from the distributions in
Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
times as many old stars in the Kepler field as young stars.
Similarly, there are \ratioobtoybplanets\ times as many old planet
hosts as young planet hosts. 
Focusing only on the tails of the distributions (0-0.3\,Gyr and
2.7-3\,Gyr), the implication is that the formation rate of stars in
the Kepler field decreased by a factor of \ratiosfr$\pm$\uncratiosfr\
over the past three billion years.

@@

Given that the history of young exoplanet discovery includes a
sociological tendency to report detailed descriptions of specific
young exoplanets, we find it necessary to briefly discuss the
properties of at least the most notable members of the set of
confirmed planets that have well-measured ages.  We first focus on
systems reported to be in the four stellar ensembles in the Kepler
field that have yielded transiting planets: Cep-Her, MELANGE-3,
Theia~520, and NGC~6811.  We then consider the broader set of Kepler
planets that are young based on the rotational and lithium-based
properties of their host stars.

@@

\paragraph{Cep-Her}
Kepler-1627Ab, Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab are
known members of the $\approx$40\,Myr old Cep-Her association
\citep{Bouma_2022a,Bouma_2022b}.  The age of various sub-components of
this association are best measured through relative isochronal
placement of the pre-main-sequence \citep[see][]{Bouma_2022b}, and
gyrochronology of stars in these clusters show them to be $<$100\,Myr
old.  The planets themselves are all on close-in orbits (5-25\,days),
with sizes from 2-4\,$R_\oplus$.  For Kepler-1627A and Kepler-1974,
the rotation, lithium, and cluster-based ages all agree within
1$\sigma$.  For Kepler-1643, while the rotation and cluster ages agree
($46^{+9}_{-7}$\,Myr; \citealt{Bouma_2022b}), the lithium-based age
($198^{+98}_{-76}$\,Myr) is 2.0$\sigma$ above the nominal cluster age
for RSG-5.  For Kepler-1975Ab (KOI\,7913Ab), the lithium age may be
similarly ``old'', though with less confidence due to the large
intrinsic scatter in K dwarf lithium calibration measurements.  For
that system, the isochronal age of its binary companion provides a
more stringent and serendipitous constraint on the $\approx$40\,Myr
age.

\paragraph{MELANGE-3}
\citet{2022AJ....164...88B} described an association in the Kepler
field that was argued to be $105\pm10$\,Myr based on a combination of
rotation, lithium, and the isochronal pre-main-sequence.  Kepler-970
and Kepler-1928 harbor two transiting mini-Neptunes inside this
association \citep{2022AJ....164...88B}.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and -1928 respectively.  The former rotation-based age
being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{2022AJ....164...88B}, where
the MELANGE-3 rotational isochrone is visually slower than
the corresponding Pleiades track.

>>>>

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
%\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}
\affiliation{51 Pegasi b Fellow}

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
  past 3\,Gyr, equally for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star formation
  history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using transits.
  %
\end{abstract}

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr
\citep[e.g.~Fig~17 of][]{Nordstrom_2004}.  This approximation clarifies the
rarity of young stars: under these assumptions, $\approx$1\% of nearby
stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
Studies of currently forming protoplanets \citep{2018A&A...617A..44K},
or of exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the Galactic star
formation rate.

Despite the observational challenges, young close-in planet discovery
has matured as a field over the past decade, primarily due to
Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and assumed planetary) ages are
precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

The main problem with the ``cluster-first'' approach is that most
nearby young stars are in the field.  Only $\approx$1\% of
stars within the nearest $\approx$500\,pc have been associated
with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing reasonable precision requirements
(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
S/N$=$3) gives $\approx$50 ``young'' transiting planets at the time
of writing.  Most of these planets are in clusters, and were
discovered by Kepler, K2, or TESS.  However, combined, those surveys
have discovered $\approx$5{,}000 planets.  Unless there is a major
bias against the discovery of sub-gigayear old planets with precisely
constrained ages, a constant star formation history implies that
an order of magnitude more young
transiting planets should be known.

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for stars in the galactic thin
disk, particularly young stars?  Second, where are the missing young
transiting planets?  Briefly, we will find that uniform is incorrect
by a factor of a few, and that the bias against the discovery of young
transiting planets is, primarily, in the measurement of
theirs ages. %, rather than in the sensitivity of the transit searches.

Given that Kepler's main mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged both isochronal ages from
an earlier analogous study \citep{Fulton_2018}, and
used a relative gyrochrone age-ranking approach that sorted stars
based on whether they rotated faster or slower than a gyrochrone age
of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
were conducted in the broader context of investigating trends in
exoplanet demographics, which also depend on stellar metallicity and
mass \citep[e.g.][]{Petigura_2018}.  These studies offered limited
sensitivity to the ages of the youngest planets, due to the nature of
isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
    positives within it;
  %
  \item Available measurements of stellar rotation periods
    in open clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    For late G dwarfs and K dwarfs, the age calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
Today, FGK star rotation-based ages are
empirically tied to the cluster age scale, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 that comprised
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}; the
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  The logarithmic scale highlights the
    dynamic range of the detections and upper limits.  Some stars with
    lithium detections do not have detected rotation periods, and
    vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

@@

Our EW measurement approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
precision of our equivalent measurement technique, after applying an
initial iteration on the \nlithiumstars\ KOIs with spectra, we
compared our lithium equivalent widths with those reported by
\citet{2018ApJ...855..115B}.  For the \nbergeroverlap\ stars in both
samples, we found broad agreement at $>$30\,m\AA, and significant
differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
required positive EWs, while we allow for negative ones.  At
$>$30\,m\AA, there is however a small offset in our respective
scales, such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  This offset is
caused by our non-treatment of the iron blend.  We therefore directly
subtract the constant mean value noted above when calculating
lithium-based ages.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
%accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
%because 91 KOIs are omitted in the 1$''$ match, 23 of which have
%either ``candidate'' or ``confirmed'' status.  Almost all of these 23
%are M dwarfs with high proper motions.  Performing the same crossmatch
%at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
%false positives, fail to yield Gaia DR3 crossmatches.}, 
%which is a
%crossmatch
of the NASA Exoplanet Archive
\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ search radius, we adopted the brightest star
as the default match.  In most such cases this was unambiguous because
there is a large brightness difference between the primary and any
apparent neighbors.  However cases with multiple stars within 4$''$
within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This is a
challenge for rotation-based ages because {\it i)} point sources
far from the main-sequence include unresolved photometric binaries,
and {\it ii}) effects other than magnetic braking influence the
rotation periods of bona fide single stars far from the main-sequence.
The added complications of reddening and metallicity further
complicate matters.  
After exploring various options, we settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.
While the details of how exactly this locus is constructed are
arbitrary and convoluted,\footnote{In practice, for K dwarfs we fitted an
$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}. We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  For F and G dwarfs, this procedure failed to produce
a reasonable locus because of shifts in the relative precision of
the isochrone ages.  We therefore visually selected KOIs that appeared
to be near the ``main sequence'', fitted a separate polynomial to
their locations, and then stitched the resulting polynomials from the
F and G dwarfs to that of the K dwarfs.} the general 
aim is to exclude stars for which there is photometric evidence
that gyrochronology may not be reliable.  The rotation period
distribution of sources excluded in this manner includes a relatively
larger fraction of ``rapid rotators'' than the main sequence sources,
likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or Figures~12 and 13
of \citet{2023ApJS..268....4F}.

@@

\subsection{Planet Quality Flags}
\label{subsec:plflags}
Some planets are more reliable than others.
We use the following additional quality indicators to assess 
the reliability and utility of a planet, and assemble them into a separate bit-flag,
$Q_{\rm planet}$.

@@

\begin{figure*}[!p]
  \begin{center}
    \leavevmode
    \subfloat{
    \includegraphics[width=\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
    }

	\vspace{-0.3cm}
    \subfloat{
    \includegraphics[width=\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Kepler's demographic cliff,} visible in the rotation-derived
    ages of stars (left) and planet hosts (middle) in the Kepler field.
    The top row corresponds to all
    stars with $T_{\rm eff}\in[3800,6200]\,$K for which we calculated rotation-based ages.
    The darker lines impose quality cuts
    on factors such as binarity, crowding, and the star's evolutionary
    state.
    The bottom row selects only stars with $T_{\rm eff}\in[4400,5400]\,$K, which
    have relatively precise ages near 2\,Gyr due to their fast spin-down
    \citep[see][]{Bouma_2023}.
    The statistical uncertainty for a mean individual star at 1, 2, and 3\,Gyr is shown, and is identical across each row.
    Finally, the right panel compares the rotation-based age distributions against star formation histories derived using CMD fitting.
    Completeness is near unity until $t$$\lesssim$3\,Gyr
    \citep{2022ApJ...937...94M}.
    \label{fig:hist_tgyro}
  }
\end{figure*}

@@

Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
stars and known planet-hosts in the Kepler field.  In detail, these
``histograms'' are sampled from the posterior probability
distributions by drawing ten random samples from each posterior
distribution, and then computing the histogram of the resulting
samples.  We truncated the plots near $4$\,Gyr since this is the
cluster age out to which our interpolation-based method is calibrated.
For Sun-like stars, the fraction of rotation signals detectable by
Kepler has been found to be near-unity up to $\approx$3\,Gyr, and it
drops to almost zero by $\approx$5\,Gyr \citep{2022ApJ...937...94M}.
This trend has the opposite functional form from our derived age
distributions.

To quantify the amount by which young stars are more rare than old, we
can group stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr as ``young'',
``intermediate-age'', and ``old''.  A simple counting exercise from
Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
times as many old stars in the Kepler field as young stars.
Similarly, there are \ratioobtoybplanets\ times as many old planet
hosts as young planet hosts.  Focusing only on the tails of the
distributions (0-0.3\,Gyr and 2.7-3\,Gyr), the implication is that the
formation rate of stars in the Kepler field decreased by a factor of
\ratiosfr$\pm$\uncratiosfr\ over the past three billion years.

@@

What are the most notable confirmed planets with well-measured ages?
These planets can be split into two classes:
those in clusters, and those with independent and consistent
rotation and lithium-based ages.

@@

To date, there are four known stellar ensembles in the Kepler field that have
yielded a total of fourteen transiting planets.
Our analysis blindly recovers each of these planets, as well as their ages.

\paragraph{Cep-Her}
The $\approx$40\,Myr old Cep-Her complex currently has its most
constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs, and it contains Kepler-1627Ab,
Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
\citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
may be older by $\approx$50\% (\citealt{Bouma_2022b};
Kerr~et.al.~in preparation).  The planets themselves are all on
close-in orbits (5-25\,days), with sizes from 2-4\,$R_\oplus$.  The
rotation, lithium, and cluster ages all agree within 1$\sigma$ for
Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the nominal cluster age for the RSG-5 subgroup.

\paragraph{MELANGE-3}
\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
the Kepler field and derived its age using a combination of rotation,
lithium, and the isochronal pre-main-sequence.  Kepler-970 and
Kepler-1928 harbor two transiting mini-Neptunes inside this
association \citep{2022AJ....164...88B}.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and -1928 respectively.  The former rotation-based age
being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.
